An Improved Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2021-08-28 simplified_sdc_intro/paper.tex
add section
some more updates

There is one additional part of the time integration algorithm.
Because $T$ evolves during the reactions (for both Strang and SDC), it
is possible for a zone to start out not in NSE but evolve into NSE
during the reaction update.  When this happens, the ODE integrator may
fail, because an excessive number of timesteps is required.  Rather
than trigger our retry mechanism and throw away the entire timestep
and start over with a smaller $\Delta t$, we first check if the zone
evolved into NSE, and if so, we redo the zone update using the NSE
prescription.  {\color{red} We optionally apply a relaxation factor, typically, $0.9$, to the NSE conditions to allow a state that is close to NSE after an integrator failure enter NSE.}

@@

@@

To solve the reaction system implicitly, the ODE solver needs the
Jacobian, $\partial \Rb/\partial \Uc^\prime$, where $\Uc^\prime =
(\rho X_k, \rho e)^\intercal$ is the subset of the conserved variables we are
integrating.  We follow the method of \cite{castro_sdc} and factor
this into two pieces,
\begin{equation}
\label{eq:factored_jac}
{\bf J} = \frac{\partial \Rb}{\partial {\bf w }} \frac{\partial {\bf w}}{\partial \Uc^\prime}.
\end{equation}
Writing this out for two species, $X_\alpha$ and $X_\beta$, we have
\begin{equation}
\Uc^\prime = \left ( \begin{array}{c} \rho X_\alpha \\ \rho X_\beta \\ \rho e \end{array} \right )
\end{equation}
For interfacing with the reaction network, we use a simpler set of variables,
\begin{equation}
{\bf w} = \left ( \begin{array}{c} X_\alpha \\  X_\beta \\ e \end{array} \right )
\end{equation}
Note: this intermediate state differs from what we used in \citet{castro_sdc} but agrees with that from \cite{strang_rnaas} in that
we use $e$ instead of $T$ as the independent variable.  If the network gives derivatives in terms
of temperature, we can convert as:
\begin{equation}
\left . \frac{\partial \phi}{\partial T} \right |_\rho =
  \left . \frac{\partial \phi}{\partial e} \right |_\rho 
  \left . \frac{\partial e}{\partial T} \right |_\rho =
  c_v \left . \frac{\partial \phi}{\partial e} \right |_\rho
\end{equation}
for some quantity $\phi(\rho, T)$.  Here we identified the specific heat
at constant volume, $c_v = \partial e/\partial T|_\rho$.

The Jacobian transformation $\partial \Uc^\prime/\partial {\bf w}$ is:
\begin{equation}
\frac{\partial \Uc^\prime}{\partial {\bf w}} = \left (
   \begin{array}{ccc}
       \rho & 0 & 0 \\
       0 & \rho & 0  \\
       \rho  e_{X_\alpha} & \rho e_{X_\beta} & \rho \\
     \end{array}\right)
\end{equation}
where we use the following notation for compactness:
\begin{equation}
e_{X_k} = \dedXd
\end{equation}
We get the inverse
(computed via SymPy, see the included Jupyter notebook) as:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial {\bf w}}{\partial \Uc^\prime} = \left (
  \begin{array}{ccc}
   \frac{1}{\rho} & 0 & 0 \\
   0 & \frac{1}{\rho} & 0 \\
    -\frac{e_{X_\alpha}}{\rho} & -\frac{e_{X_\beta}}{\rho} & \frac{1}{\rho} \\
   \end{array}\right)
\renewcommand{\arraystretch}{1}
\end{equation}

The reaction vector is
\begin{equation}
\Rb(\Uc^\prime) = \left (  \begin{array}{c} \rho \omegadot_\alpha \\ \rho \omegadot_\beta \\ \rho \Sdot \end{array} \right )
\end{equation}
We take all the quantities to be functions of $\rho$, $e$, and $X_k$,
but since $\rho$ doesn't change from the reactions (the reactive
source of the continuity equation is zero), it is held constant in
these derivatives.  The Jacobian is computed as $\partial \Rb/\partial
{\bf w}$:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial \Rb}{\partial {\bf w}} = \left (
  \begin{array}{ccc}
     \rho \frac{\partial \omegadot_\alpha}{\partial X_\alpha} &
     \rho \frac{\partial \omegadot_\alpha}{\partial X_\beta} & 
     \rho \frac{\partial \omegadot_\alpha}{\partial e} \\
      %
     \rho \frac{\partial \omegadot_\beta}{\partial X_\alpha} &
     \rho \frac{\partial \omegadot_\beta}{\partial X_\beta} &
     \rho \frac{\partial \omegadot_\beta}{\partial e} \\
     %
     \rho \frac{\partial \Sdot}{\partial X_\alpha} &
     \rho \frac{\partial \Sdot}{\partial X_\beta} &
     \rho \frac{\partial \Sdot}{\partial e} \\
  \end{array}
  \right )
\renewcommand{\arraystretch}{1}
\end{equation}

While VODE can compute the entire Jacobian, ${\bf J}$ numerically via differencing, we found
that does not give reliable results.  Instead, we compute $\partial {\bf w}/\partial {\bf \Uc}^\prime$ analytically and $\partial {\bf R}/\partial {\bf w}$ numerically via one-sided differencing, following
the algorithm in \cite{lsode} to minimize numerical noise.
The final Jacobian is found by multiplying these two:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial \Rb}{\partial \Uc^\prime} = \left (
  \begin{array}{ccc}
    \frac{\partial \omegadot_\alpha}{\partial X_\alpha} -\frac{\partial \omegadot_\alpha}{\partial e} e_{X_\alpha} &
    \frac{\partial \omegadot_\alpha}{\partial X_\beta} -\frac{\partial \omegadot_\alpha}{\partial e}  e_{X_\beta} &
    \frac{\partial \omegadot_\alpha}{\partial e} \\
     %
    \frac{\partial \omegadot_\beta}{\partial X_\alpha} -\frac{\partial \omegadot_\beta}{\partial e} e_{X_\alpha} &
    \frac{\partial \omegadot_\beta}{\partial X_\beta} -\frac{\partial \omegadot_\beta}{\partial e}  e_{X_\beta} &
    \frac{\partial \omegadot_\beta}{\partial e} \\
     %
     \frac{\partial \Sdot}{\partial X_\alpha} - \frac{\partial \Sdot}{\partial e} e_{X_\alpha} &
     \frac{\partial \Sdot}{\partial X_\beta} - \frac{\partial \Sdot}{\partial e} e_{X_\beta} &
     \frac{\partial \Sdot}{\partial e} \\
  \end{array}
  \right )
\renewcommand{\arraystretch}{1}
\end{equation}

>>>>

There is one additional part of the time integration algorithm.
Because $T$ evolves during the reactions (for both Strang and SDC), it
is possible for a zone to start out not in NSE but evolve into NSE
during the reaction update.  When this happens, the ODE integrator may
fail, because an excessive number of timesteps is required.  Rather
than trigger our retry mechanism and throw away the entire timestep
and start over with a smaller $\Delta t$, we first check if the zone
evolved into NSE, and if so, we redo the zone update using the NSE
prescription.  To ensure that this is not triggered at the very start
of integration, we require a minimum number of steps (10) to have
elapsed before checking for NSE.  After we leave the integrator, the
state is returned to the burner driver where we finish the integration
via NSE if the conditions satisfy the NSE criteria.  We optionally
apply a relaxation factor, typically, $0.9$, to the NSE conditions to
allow a state that is close to NSE after an integrator failure enter
NSE.

@@

\subsection{Reacting Buoyant Bubble}

We next consider a buoyant, reacting bubble in a stratified,
plane-parallel atmosphere.  This test exercises the treatment of
hydrodynamical sources in the SDC integration.  We use the set of
initial conditions that was first shown in \cite{ABNZ:III}.

@@

To solve the reaction system implicitly, the ODE solver needs the
Jacobian, $\partial \Rb/\partial \Uc^\prime$, where $\Uc^\prime =
(\rho X_k, \rho e)^\intercal$ is the subset of the conserved variables we are
integrating.  We follow the method of \cite{castro_sdc} and factor
this into two pieces,
\begin{equation}
\label{eq:factored_jac}
{\bf J} = \frac{\partial \Rb}{\partial {\bf w }} \frac{\partial {\bf w}}{\partial \Uc^\prime}.
\end{equation}
where the state ${\bf w}$ is chosen to match the set of variables used to evaluate the reaction rates.
Writing this out for two species, $X_\alpha$ and $X_\beta$, we have
\begin{equation}
\Uc^\prime = \left ( \begin{array}{c} \rho X_\alpha \\ \rho X_\beta \\ \rho e \end{array} \right )
\end{equation}
For interfacing with the reaction network, we use
\begin{equation}
{\bf w} = \left ( \begin{array}{c} X_\alpha \\  X_\beta \\ T \end{array} \right )
\end{equation}
Note: even though we are using $T$ here instead of $e$, we still do the overall ODE integration
in terms of $(\rho e)$, consistent with the Strang method described in \cite{strang_rnaas}.
%%  in that
%% we use $e$ instead of $T$ as the independent variable.  If the network gives derivatives in terms
%% of temperature, we can convert as:
%% \begin{equation}
%% \left . \frac{\partial \phi}{\partial T} \right |_\rho =
%%   \left . \frac{\partial \phi}{\partial e} \right |_\rho 
%%   \left . \frac{\partial e}{\partial T} \right |_\rho =
%%   c_v \left . \frac{\partial \phi}{\partial e} \right |_\rho
%% \end{equation}
%% for some quantity $\phi(\rho, T)$.  Here we identified the specific heat
%% at constant volume, $c_v = \partial e/\partial T|_\rho$.

The Jacobian transformation $\partial \Uc^\prime/\partial {\bf w}$ is:
\begin{equation}
\frac{\partial \Uc^\prime}{\partial {\bf w}} = \left (
   \begin{array}{ccc}
       \rho & 0 & 0 \\
       0 & \rho & 0  \\
       \rho  e_{X_\alpha} & \rho e_{X_\beta} & \rho c_v \\
     \end{array}\right)
\end{equation}
where we use the following notation for compactness:
\begin{equation}
e_{X_k} = \dedXd
\end{equation}
and the specific heat at constant volume is
\begin{equation}
c_v = \left . \frac{\partial e}{\partial T} \right |_{\rho, X_k}
\end{equation}
We get the inverse
(computed via SymPy) as:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial {\bf w}}{\partial \Uc^\prime} = \left (
  \begin{array}{ccc}
   \frac{1}{\rho} & 0 & 0 \\
   0 & \frac{1}{\rho} & 0 \\
    -\frac{e_{X_\alpha}}{\rho c_v} & -\frac{e_{X_\beta}}{\rho c_v} & \frac{1}{\rho c_v} \\
   \end{array}\right)
\renewcommand{\arraystretch}{1}
\end{equation}

The reaction vector is
\begin{equation}
\Rb(\Uc^\prime) = \left (  \begin{array}{c} \rho \omegadot_\alpha \\ \rho \omegadot_\beta \\ \rho \Sdot \end{array} \right )
\end{equation}
We take all the quantities to be functions of $\rho$, $e$, and $X_k$,
but since $\rho$ doesn't change from the reactions (the reactive
source of the continuity equation is zero), it is held constant in
these derivatives.  The Jacobian is computed as $\partial \Rb/\partial
{\bf w}$:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial \Rb}{\partial {\bf w}} = \left (
  \begin{array}{ccc}
     \rho \frac{\partial \omegadot_\alpha}{\partial X_\alpha} &
     \rho \frac{\partial \omegadot_\alpha}{\partial X_\beta} & 
     \rho \frac{\partial \omegadot_\alpha}{\partial T} \\
      %
     \rho \frac{\partial \omegadot_\beta}{\partial X_\alpha} &
     \rho \frac{\partial \omegadot_\beta}{\partial X_\beta} &
     \rho \frac{\partial \omegadot_\beta}{\partial T} \\
     %
     \rho \frac{\partial \Sdot}{\partial X_\alpha} &
     \rho \frac{\partial \Sdot}{\partial X_\beta} &
     \rho \frac{\partial \Sdot}{\partial T} \\
  \end{array}
  \right )
\renewcommand{\arraystretch}{1}
\end{equation}

The final Jacobian is found by multiplying these two:
\begin{equation}
\renewcommand{\arraystretch}{1.5}
\frac{\partial \Rb}{\partial \Uc^\prime} = \left (
  \begin{array}{ccc}
    \frac{\partial \omegadot_\alpha}{\partial X_\alpha} - \frac{e_{X_\alpha}}{c_v} \frac{\partial \omegadot_\alpha}{\partial T} &
    \frac{\partial \omegadot_\alpha}{\partial X_\beta} - \frac{e_{X_\beta}}{c_v} \frac{\partial \omegadot_\alpha}{\partial T} &
    \frac{1}{c_v} \frac{\partial \omegadot_\alpha}{\partial T} \\
     %
    \frac{\partial \omegadot_\beta}{\partial X_\alpha} - \frac{e_{X_\alpha}}{c_v} \frac{\partial \omegadot_\beta}{\partial T} &
    \frac{\partial \omegadot_\beta}{\partial X_\beta} - \frac{e_{X_\beta}}{c_v} \frac{\partial \omegadot_\beta}{\partial T} &
    \frac{1}{c_v} \frac{\partial \omegadot_\beta}{\partial T} \\
     %
     \frac{\partial \Sdot}{\partial X_\alpha} -  \frac{e_{X_\alpha}}{c_v} \frac{\partial \Sdot}{\partial T} &
     \frac{\partial \Sdot}{\partial X_\beta} -  \frac{e_{X_\beta}}{c_v} \frac{\partial \Sdot}{\partial T} &
     \frac{1}{c_v} \frac{\partial \Sdot}{\partial T} \\
  \end{array}
  \right )
\renewcommand{\arraystretch}{1}
\end{equation}
We note that the form of these entries is the same as one would arrive at if you start with the rates expressed as
$\omega_k(\rho, T(\rho, X_j, e), X_j)$ and note that constant $e$ implies that
\begin{equation}
\left . \frac{\partial T}{\partial X_k} \right |_{\rho, e} = - \frac{e_{X_k}}{c_v}
\end{equation}

While VODE can compute the entire Jacobian, ${\bf J}$ numerically via
differencing, we found that does not give reliable results.  Instead,
we compute compute the derivatives with respect to $X_k$ and $T$
one-sided differencing, following the algorithm in \cite{lsode} to
minimize numerical noise.  We then use the equation of state to
compute $e_{X_k}$ and $c_v$ and construct the entries of the final
Jacobian.
